Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-02 analysis/paper/paper.qmd
edited and added introduction

With the introduction of new technologies, the Korean Paleolithic transitioned from the Early to the Late Paleolithic periods. The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Another notable change is the selective use of raw materials along with the emergence of new tools. Previously quartzite and vein quartz were the most commonly used for core and flake tools but finer grained materials such as silicified tuff (shale), chert, hornfels, and obsidian became more important to the lithic technology during the Late Paleolithic. While people still used coarse materials with existing tools, they selectively chose finer materials for newly introduced tools. [@Seong_2003; @seong2004quartzite]. Stemmed points are considered the first evidence of the new technology, as well as the Late Paleolithic, since they led to the whole set of changes [@seong2016eastern; @seong2008tanged]. Due to the fact that stemmed points originated in Korea in Northeast Asia and are closely associated with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies have examined their use, whereas researchers often discuss stemmed points relating to their origin and chronology of the Korean Late Paleolithic and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

>>>>

With the introduction of new technologies, the Korean Paleolithic transitioned from the Early to the Late Paleolithic periods. The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern]. Another notable change is the selective use of raw materials along with the emergence of new tools. Previously quartzite and vein quartz were the most commonly used for core and flake tools but finer grained materials such as silicified tuff (shale), chert, hornfels, and obsidian became more important to the lithic technology during the Late Paleolithic. While people still used coarse materials with existing tools, they selectively chose finer materials for newly introduced tools. [@Seong_2003; @seong2004quartzite]. Stemmed points are considered the first evidence of the new technology, as well as the Late Paleolithic, since they led to the whole set of changes [@seong2016eastern; @seong2008tanged]. This is related to the fact that stemmed points originate from Korea in Northeast Asia and have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did]. Despite the importance of stemmed points, only a few studies have examined their use, whereas researchers often discuss stemmed points relating to their origin and chronology of the Korean Late Paleolithic and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

To understand the role stemmed points played in the technological transition, this study examines the functions of stemmed points. We use the tip cross-sectional area (TCSA) metric to calculate artifact function and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different functions [@lombard2021variation]. We then explore the relationship between and function raw materials, artifact size and discard location, and how these change over time. Our main questions are: What were stemmed points used for? How diverse were their functions? What are the temporal patterns in stemmed point functions? As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period. In order to understand how a certain weapon-tip type was chosen, we premise that human groups had the cognitive capacity to choose proper weapons depending on their hunting strategies and socio-environmental circumstances.
